---
title: "Data 608 Module 3"
author: "Leticia Salazar"
date: "10-09-2022"
output:
  html_document: 
    theme: united
    highlight: kate
  html_notebook: default
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
$~$

### Mortality from all 50 states and the District of Colombia

I have provided you with data about mortality from all 50 states and the District of Columbia. Please access it at https://github.com/charleyferrari/CUNY_DATA608/tree/master/module3/data You are invited to gather more data from our provider, the CDC WONDER system, at https://wonder.cdc.gov/ucd-icd10.html.

* This assignment must be done in R. It must be done using the ‘shiny’ package.

* It is recommended you use an R package that supports interactive graphing such as plotly, or vegalite, but this is not required.

* Your apps must be deployed, I won’t be accepting raw files. Luckily, you can pretty easily deploy apps with a free account at shinyapps.io



$~$

#### Load Libraries
```{r, load libraries, warning=FALSE, message=FALSE}
library(shiny)
library(tidyverse)
library(ggplot2)
library(rsconnect)
library(plotly)
library(shinydashboard)
```


#### Load Data
```{r cars}
data <- read.csv("https://raw.githubusercontent.com/charleyferrari/CUNY_DATA_608/master/module3/data/cleaned-cdc-mortality-1999-2010-2.csv", header = TRUE)
```


```{r}
str(data)
```


```{r}
summary(data)
```


```{r}
colnames(data)
```


```{r}
colSums(is.na(data))
```


```{r, authorizing account, echo=FALSE}
rsconnect::setAccountInfo(name='letisalba', token='18C2DF783293E23721636AD4A43F97C6', secret='seSfGaZ7eiF653BaFtyAGHFls/Z66OE/t7Sk2b9A')
```


-------------
### Question 1:
As a researcher, you frequently compare mortality rates from particular causes across different States. You need a visualization that will let you see (for 2010 only) the crude mortality rate, across all States, from one cause (for example, Neoplasms, which are effectively cancers). Create a visualization that allows you to rank States by crude mortality for each cause of death.

```{r question 1}
# Filtering data
infec_diseases <- filter(data, Year == '2010' & ICD.Chapter == 'Certain infectious and parasitic diseases')
infec_diseases <- infec_diseases %>% 
  arrange(Crude.Rate)
head(infec_diseases)


# creating barplot
fig <- infec_diseases %>% plot_ly()
fig <- fig %>% add_trace(x = ~Crude.Rate, y = ~State, type = 'bar',
             text = 'y', textposition = 'auto',
             marker = list(color = 'rgb(158,202,225)',
                           line = list(color = 'rgb(8,48,107)', width = 1.5)))
fig <- fig %>% layout(title = "Certain infectious and parasitic diseases",
         barmode = 'group',
         xaxis = list(title = "Crude rate"),
         yaxis = list(title = "States"))

#Shiny app prep
# Define UI for app 
ui <- fluidPage(
  # App title ----
  titlePanel("Crude Mortality Rate"),
  # Sidebar layout with input and output definitions ----
  sidebarLayout(
    # Sidebar panel for inputs ----
    sidebarPanel(
      # Input: 
      selectInput(inputId = 'y','Selected ICD.Chapter:',
                  choices = infec_diseases$ICD.Chapter,
                  selected = 'State')

    ),

    # Main panel for displaying outputs ----
    mainPanel(
      # Output: Create barchart
      tabsetPanel(type = "tabs",
                  tabPanel("Plot", plotlyOutput('distPlot'))
     )

    )
  )
)
# Define server logic required to draw a bargraph
server <- function(input, output) {
  output$distPlot <- renderPlotly({fig})
}

# Run the application
shinyApp(ui, server)
```


$~$

-------------

### Question 2:
Often you are asked whether particular States are improving their mortality rates (per cause) faster than, or slower than, the national average. Create a visualization that lets your clients see this for themselves for one cause of death at the time. Keep in mind that the national average should be weighted by the national population.


```{r question 2}
# Filter data
infec_diseases_2 <- data %>% 
  group_by(Year, ICD.Chapter) %>% 
  filter(ICD.Chapter == "Certain infectious and parasitic diseases") %>% 
  mutate(Crude.Rate_2 = round(
    sum(Deaths) / sum(Population) * 100000),3) %>%
  group_by(Year, ICD.Chapter, State)

# Creating plot

# This plot was not working out for some reason
# state = "CA"
# fig_2<- infec_diseases_2 %>%
#   plot_ly(x=~Year, y=~Crude.Rate, type ='bar',
#           #text = y, textposition = 'auto',
#           marker = list(color = 'rgb(158,202,225)'), 
#           name = 'State') %>% # Chart State and US Crude Rates next to one another
#   add_trace(x=~Year, y=~Crude.Rate_2, type='bar',
#             #text = y, textposition = 'auto',
#             marker = list(color = 'rgb(58,200,225)'), name = 'US')  %>%
#  layout(title = "Crude death rate by state [CA]",
#         barmode = 'group',
#         xaxis = list(title = "Year"),
#         yaxis = list(title = "Crude rate"))
# fig_2


# Creating second plot
# Trying this new plot and it came out much better than plot 1
state = "CA"
fig_2 <- infec_diseases_2 %>%
  as_tibble()  %>% 
  filter(., State == state) %>% 
  select(., Year, Crude.Rate, Crude.Rate_2) %>%
  plot_ly(x = ~Year, y = ~Crude.Rate, type='bar',
          text = ~Crude.Rate, textposition = 'auto',
          marker = list(color = 'rgb(158,202,225)'), 
          name = 'State') %>% 
  add_trace(x = ~Year, y = ~Crude.Rate_2, type='bar',        
            text = ~Crude.Rate_2, textposition = 'auto',
            marker = list(color = 'rgb(58,200,225)'), name = 'US')  %>%
 layout(title = "Crude death rate by state [CA]",
        barmode = 'group',
        xaxis = list(title = "Year"),
        yaxis = list(title = "Crude rate"))


# Shiny app prep
# Define UI for app 
ui <- fluidPage(
  # App title ----
  titlePanel("Crude Mortality Rate"),
  # Sidebar layout with input and output definitions ----
  sidebarLayout(
    # Sidebar panel for inputs ----
    sidebarPanel(
      # Input: 
      selectInput(inputId = 'y','Selected ICD.Chapter:',
                  choices = infec_diseases_2$ICD.Chapter,
                  selected = 'State')

    ),

    # Main panel for displaying outputs ----
    mainPanel(
      # Output: Create barchart
      tabsetPanel(type = "tabs",
                  tabPanel("Plot", plotlyOutput('distPlot2'))
     )

    )
  )
)
server <- function(input, output) {
  output$distPlot2 <- renderPlotly({fig_2})
}

# Run the application
shinyApp(ui, server)
```


```{r, echo=FALSE}
#rsconnect::deployApp('/Users/letiix3/Desktop/DATA_608/module3')
```


#### References:
* [Plotly](https://plotly.com/r/bar-charts/)

