---
title: "Data 608 - Final Project"
author: "Leticia Salazar"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
    highlight: kate
  html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::include_graphics("")
```

# Polycystic Ovarian Syndrome (PCOS)

## Overview: 

Polycystic ovarian syndrome (PCOS) is a reproductive hormonal imbalance condition. It's believed that genetics and environmental factors can cause PCOS that affect the body physically and emotionally along with the metabolism, overall health and appearance in women. PCOS causes problems in ovaries making it hard for women to have a healthy menstrual cycle leading to the development of cysts and infertility. Although very common in women of reproductive age, PCOS may begin shortly after puberty but can also develop during the later teenage years and early adulthood.

$~$

Hormones that are involved in PCOS are:

* **Androgens: ** aka "male hormones" are present in women with PCOS at higher levels than usual. Excess in androgens can cause symptoms such as acne, unwanted hair, thinning hair, and irregular periods.

* **Insulin: ** allows the body to absorb glucose (blood sugar) into the cells for energy. In PCOS, the body doesnt respond to insulin as intended therefore, evelations in blood glucose levels can be assessed. Such evelations then lead to increased production of androgen.

* **Progesterone: ** vital hormone for menstruation and pregnancy; lack of progesterone contributes to irregular periods.

$~$

PCOS Symptoms:

Many of these symptoms can be attributed to other causes or go unnoticed but it is very common for PCOS to go undiagnosed for some time. Here are some symptoms that help with the diagnosis:

* **Irregular periods: ** 

* **Polycystic ovaries: ** 

* **Excess androgen: ** 

$~$

Other symptoms may include:

* **Weight gain: ** 

* **Fatigue: ** 

* **Unwanted hair growth: ** 

* **Thinning hair on the head: ** 

* **Infertility: ** 

* **Acne: ** 

* **Darkening of skin: ** 

* **Mood changes: ** 

* **Pelvic pain: ** 

* **Headaches: ** 

* **Sleep problems: ** 

* **Depression: ** 

*It's good to note that not everyone who is diagnosed with PCOS experiences all of these symptoms and should always consult with a their PCP or OBGYN to get an accurate diagnosis.*


### Libraries used:
```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(corrplot)
```


### Load data:
```{r}
pcos <- read.csv("https://raw.githubusercontent.com/letisalba/DATA_608/master/Final%20Project/csv/PCOS_infertility.csv")
pcos_infertility <- read.csv("https://raw.githubusercontent.com/letisalba/DATA_608/master/Final%20Project/csv/PCOSData_without_infertility.csv")
```


### Data Exploration:

The first data set `pcos` includes 541 entries and a total of 6 columns. The second data set `pcos_infertility` includes 541 entires and a total of 45 columns

```{r}
glimpse(pcos)
```

```{r}
glimpse(pcos_infertility)
```




## Project Plan:

* Merge the two files into one
* Deal with missing values (such as BMI, FSH.LH, Waist.Hip.Ratio)


$~$

## References:

* Stewart, M. M., &amp; Foster, S. (2012). PCOS awareness association. PCOS Awareness Association. Retrieved October 9, 2022, from https://www.pcosaa.org/ 
* Bartlett, E., &amp; Erlich, L. (2015). Feed your fertility: Your guide to cultivating a healthy pregnancy with traditional Chinese medicine, real food, and holistic living. Fair Winds Press. 